Explore tumor cell classification with CellAssign

### 2024-05-07 analyses/ewings-cell-type/02-cellassign-tumor-cell-classification.Rmd
add heatmaps to notebook

@@

```{r}
# function for creating a binary matrix with cell types as columns and marker genes as rows 
build_binary_mtx <- function(marker_genes_df,
                             gene_col, 
                             cell_type_col) {

  binary_mtx <- marker_genes_df |>
  dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
  unique() |> 
  tidyr::pivot_wider(
    id_cols = {{gene_col}},
    names_from = {{cell_type_col}},
    values_from = {{cell_type_col}},
    values_fn = length,
    values_fill = 0
  ) |>
  # add a column with no marker genes
  # cell assign will assign cells to "other" when no other cell types are appropriate
  dplyr::mutate(other = 0)

  return(binary_mtx)
}

@@

```{r}
tumor_ref_mtx <- build_binary_mtx(tumor_markers_df, 
                                  gene_col = "ensembl_gene_id",
                                  cell_type_col = "cell_type")

@@

@@

@@

```{r}
# first read in tumor normal classifications 
classifications_df <- readr::read_tsv(classifications_file)

@@

celltype_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      ) |> 
  # join with tumor normal
  dplyr::left_join(classifications_df) |> 
  # join with filtered cell type assignments from cellassign 
  dplyr::left_join(filtered_celltype_assignments, by = c("barcodes" = "barcode"))

@@

```{r}
# build the reference matrix 
visser_ref_mtx <- build_binary_mtx(all_markers_df, 
                                   gene_col = "ensembl_gene_id",
                                   cell_type_col = "cell_type")

@@

table(visser_celltype_assignments$celltype)
```
Here we see that there are still only a few hundred tumor cells and most cells are endothelial cells. 
Generally, I expect that there to be a large portion of tumor cells with much fewer cells from normal cell types.

@@

```{r}
# build the reference matrix 
panglao_tumor_ref_mtx <- build_binary_mtx(panglao_tumor_markers_df, 
                                          gene_col = "ensembl_gene_id",
                                          cell_type_col = "cell_type")

@@

>>>>

```{r}
# function for creating a binary matrix with cell types as columns and marker genes as rows 
build_binary_df <- function(marker_genes_df,
                            gene_col, 
                            cell_type_col) {

  binary_df <- marker_genes_df |>
    dplyr::select({{cell_type_col}}, {{gene_col}}) |> 
    unique() |> 
    tidyr::pivot_wider(
      id_cols = {{gene_col}},
      names_from = {{cell_type_col}},
      values_from = {{cell_type_col}},
      values_fn = length,
      values_fill = 0
    ) |>
    # add a column with no marker genes
    # cell assign will assign cells to "other" when no other cell types are appropriate
    dplyr::mutate(other = 0)

  return(binary_df)
}

@@

@@

```{r}
tumor_ref_mtx <- build_binary_df(tumor_markers_df, 
                                  gene_col = "ensembl_gene_id",
                                  cell_type_col = "cell_type")

@@

```{r}
# heatmap of prediction scores 
tumor_predictions |>
  dplyr::select(-barcode) |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE)
```

@@

```{r}
# heatmap of prediction scores 
filtered_predictions |>
  dplyr::select(-barcode) |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE)
```

@@

```{r}
# first read in tumor normal classifications 
# obtained from manual classification in `01-marker-gene-tumor-classifications.Rmd`
classifications_df <- readr::read_tsv(classifications_file)

@@

celltype_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # join with tumor normal
  dplyr::left_join(classifications_df, by = "barcodes") |> 
  # join with filtered cell type assignments from cellassign 
  dplyr::left_join(filtered_celltype_assignments, by = c("barcodes" = "barcode"))

@@

```{r}
# build the reference matrix 
visser_ref_mtx <- build_binary_df(all_markers_df, 
                                   gene_col = "ensembl_gene_id",
                                   cell_type_col = "cell_type")

@@

table(visser_celltype_assignments$celltype)
```

```{r}
# heatmap of prediction scores 
visser_predictions |>
  dplyr::select(-barcode) |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE)
```

Here we see that there are still only a few hundred tumor cells and most cells are endothelial cells. 
Generally, I expect that there to be a large portion of tumor cells with much fewer cells from normal cell types.

@@

```{r}
# build the reference matrix 
panglao_tumor_ref_mtx <- build_binary_df(panglao_tumor_markers_df, 
                                          gene_col = "ensembl_gene_id",
                                          cell_type_col = "cell_type")

@@

```{r}
# heatmap of prediction scores 
panglao_predictions |>
  dplyr::select(-barcode) |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE)
```
